Tropical Mathematics for the Lambda Calculus

### 2023-01-16 section3C.tex
commit

A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see [Giulio,?] for details.
In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:

@@

There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the \lamcalc Taylor expansion.
This is expressed by the following property (see [Definition 4.22, Giulio], where it is called the fact of ``modeling the Taylor expansion'').

@@

This property is interesting because of [Theorem 4.23,Giulio], which says that, then, in the model the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$.

>>>>

A model of the $\STDLC$ is usually understood as so-called \emph{Cartesian closed differential categories} (CC$\partial$C), see \cite{Manzo2010} for details.
In order to treat the $+$ and the constructor $D[\_]\cdot (\_)$ of $\STDLC$, the main features of a CC$\partial$C $\C C$ are that:

@@

There is more: not only $\LREL_!$ is a model of the $\STDLC$, but this model is also well-behaved w.r.t.\ to the \lamcalc Taylor expansion.
This is expressed by the following property (see [Definition 4.22, \cite{Manzo2010}], where it is called the fact of ``modeling the Taylor expansion'').

@@

This property is interesting because of [Theorem 4.23, \cite{Manzo2010}], which says that, then, in the model the series corresponding to the Taylor expansion of an ordinary $\lam$-term $M$ converges to $M$.

### 2023-01-16 section3intro.tex
commit

For a fixed \emph{continuous} semi-ring $Q$ [Def. II.5, Weighted], the well-known \emph{quantitative relational model}, is the category $\QREL$ of sets and set-indexed matrices with coefficients in $Q$ ([Weighted] would call it $Q^\Pi$).
We must fix a convention for matrices: following [Weighted, MonTop, Ehrhard], we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$.
Instead, we prefer to see a $t\in\HOM{\QREL}{X}{Y}$ as giving rise to a map $\hat t:Q^X\to Q^Y$ defined by $\hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a$.
In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.

\begin{remark}
 The category $\QREL$ is (equivalent to) a subcategory of the category $Q\SF{Mod}$ of \emph{complete} $Q$-semimodules.
 If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
 We take this viewpoint in Section~\ref{sec:QMod}.
\end{remark}

The assumption of $Q$ being continuous (which can be weakened, see [??]) is used in order to make the potentially infinite sum, appearing in the formula for the composition, converge.

@@

Since $\Lawv$ is a continuous (commutative) semiring, [Proposition III.3, Weighted] immediately applies and gives:

@@

Unwrapping [Definition II.9, Weighted], this means that:
$\HOM{\LREL}{X}{Y}$ is a continuous $\Lawv$-semimodule, with semimodule operations defined pointwise;
$\LREL$ is a continuous $\Lawv$-category, i.e.\ composition of morphisms commutes with $\inf$'s;
$\LREL$ is linear, i.e.\ pre- and post-composition with any morphism in any $\HOM{\LREL}{X}{Y}$ are automorphisms on it.

>>>>

For a fixed \emph{continuous} semi-ring $Q$ [Def. II.5, \cite{Manzo2013}], the well-known \emph{quantitative relational model}, is the category $\QREL$ of sets and set-indexed matrices with coefficients in $Q$ (\cite{Manzo2013} would call it $Q^\Pi$).
We must fix a convention for matrices: following \cite{Manzo2013, Hofmann2014, Ehrhard2005}, we fix $\HOM{\QREL}{X}{Y}:=Q^{X\times Y}$ with composition $st:X\times Z\to Q$ of $s:Y\times Z\to Q$ and $t:X\times Y\to Q$ defined by $(st)_{a,c}:=\sum\limits_{b\in Y} s_{b,c}t_{a,b}$.
In linear algebra, a map $X\times Y\to Q$ is usually called a ``$Y\times X$-matrix'', meaning $Y$ rows and $X$ columns.
In particular, the product of such a matrix for a vector defines a map $Q^Y\to Q^X$.
Instead, we prefer to see a $t\in\HOM{\QREL}{X}{Y}$ as giving rise to a map $\hat t:Q^X\to Q^Y$ defined by $\hat t(x)_a:=\sum\limits_{b\in Y} t_{a,b}x_a$.
In $\QREL^{op}$ (which corresponds to systematically taking transpose matrices), composition coincides with the product matrix/matrix and $\hat{(\cdot)}$ with the product matrix/vector.
In order to avoid confusion, we will refer to a $t\in Q^{X\times Y}$ just as a \emph{matrix from $X$ to $Y$}.
As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.

\begin{remark}
 The category $\QREL$ is (equivalent to) a subcategory of the category $Q\SF{Mod}$ of \emph{complete} $Q$-semimodules.
 If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
 We take this viewpoint in Section~\ref{section6}.
\end{remark}

The assumption of $Q$ being continuous (which can be weakened, see \cite{???}) is used in order to make the potentially infinite sum, appearing in the formula for the composition, converge.

@@

Since $\Lawv$ is a continuous (commutative) semiring, [Proposition III.3, \cite{Manzo2013}] immediately applies and gives:

@@

Unwrapping [Definition II.9, \cite{Manzo2013}], this means that:
$\HOM{\LREL}{X}{Y}$ is a continuous $\Lawv$-semimodule, with semimodule operations defined pointwise;
$\LREL$ is a continuous $\Lawv$-category, i.e.\ composition of morphisms commutes with $\inf$'s;
$\LREL$ is linear, i.e.\ pre- and post-composition with any morphism in any $\HOM{\LREL}{X}{Y}$ are automorphisms on it.

### 2023-01-16 section5.tex
commit

In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs. 
The general pattern is to consider a program $M$, which can be expressed in the form of a polynomial or a power series, and 
investigate what quantitative properties of $M$ can be deduced from properties of its tropicalization $\trop M$ (which are often easier to study).
By the way, this is the approach taken in several well-studied applications of tropical geometry (e.g.~for optimization problems related to machine learning \cite{}).

@@

The possibility of using the relational model over the tropical semi-ring for ``best case'' resource analysis has already been explored in \cite{Manzo}. Notably, they considered an interpretation of a language for $\B{PCF}$ with non-deterministic choice in which each $\lambda$-abstraction and each occurrence of the fixpoint operator $Y$ is assigned a ``weight'' 1, and showed that for any program $M$ of type $\B{nat}$, 
the value of the interpretation $\model{M}\in \Lawv^{\BB N}$ on a particular natural number $k$, i.e.~$\model{M}(k)\in \Lawv$, corresponds to the \emph{minimum} number of $\beta$- or $\TT{fix}$-redexes reduced in a reductions sequence from $M$ to $\underline n$. 
In the next paragraph we will illustrate an analogous ``best case'' analysis for probabilistic programs.

@@

Secondly, the Lipschitz conditions studied in the previous section allow us to reason on program distances in a \emph{compositional} way: suppose, as before, that $M,N:A$ are two programs such that $M$ improves on $N$ by $\epsilon$, and let $\TT C[-]:A \to \B{nat}$ indicate a context; knowing that the interpretation of $\TT C$ is $k$-Lipschitz-continuous on some open set containing both $\model M$ and $\model N$, allows us to immediately deduce that $\TT C[M]$ improves on $\TT C[N]$ by $k \epsilon$. 
Observe that this will typically be the case when the Taylor expansions of $\TT C[M]$ and $\TT C[N]$ actually yields a \emph{finite} sum of at most $k$ terms, i.e.~when 
\begin{align}
\TT C[M] = \sum_{i=0}^{k} \TT D^{(k)}\Big[\lambda x.\TT C[x]\Big]\cdot M^{k}
\end{align}
and similarly for $\TT C[N]$. It is here worth recalling that, for $\STDLC$, a well-known result \cite{} is that the Taylor expansion of a closed application $MN$ is always finite, although its non-zero coefficients may be arbitrarily high. 
Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{}, 
a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
We mention this point in the Section~\ref{sec:concl}

@@

\begin{fact}
 By using [Weighted], it can be shown that $\LREL_!$ is a model of such calculus $\STLC_\oplus$. {\color{red}GIUSTIFICARLO BENE ?}
\end{fact}

@@

More formally, differential privacy is defined as follows \cite{}:
\begin{definition}
Let $f: \mathsf{db}\to [0,1]^{X}$ and $\epsilon \in \BB R_{\geq 0}$. $f$ is said \emph{$\epsilon$-differentially private} if for all $x,x'\in \mathsf{db}$
differing by $L$ items, for all $s\in X$, 
$$
f(x)(s) \ = \ e^{\epsilon L} f(x')(s)
$$
\end{definition}

A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{}. Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.

The use of tropical semantics suggests how resource analysis could also be used to provide bounds for differential privacy. 
Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{}). Then, if we discover, either by studying differential properties of $f$, or using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{??}), 
 that the tropicalization $\trop f$ satisfies a Lipschitz condition, we may use this fact to deduce that $f$ is differentially private, as shown by the result below.

>>>>

In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs. 
The general pattern is to consider a program $M$, which can be expressed in the form of a polynomial or a power series, and 
investigate what quantitative properties of $M$ can be deduced from properties of its tropicalization $\trop M$ (which are often easier to study).
By the way, this is the approach taken in several well-studied applications of tropical geometry (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}).

@@

The possibility of using the relational model over the tropical semi-ring for ``best case'' resource analysis has already been explored in \cite{Manzo2013}. Notably, they considered an interpretation of a language for $\B{PCF}$ with non-deterministic choice in which each $\lambda$-abstraction and each occurrence of the fixpoint operator $Y$ is assigned a ``weight'' 1, and showed that for any program $M$ of type $\B{nat}$, 
the value of the interpretation $\model{M}\in \Lawv^{\BB N}$ on a particular natural number $k$, i.e.~$\model{M}(k)\in \Lawv$, corresponds to the \emph{minimum} number of $\beta$- or $\TT{fix}$-redexes reduced in a reductions sequence from $M$ to $\underline n$. 
In the next paragraph we will illustrate an analogous ``best case'' analysis for probabilistic programs.

@@

Secondly, the Lipschitz conditions studied in the previous section allow us to reason on program distances in a \emph{compositional} way: suppose, as before, that $M,N:A$ are two programs such that $M$ improves on $N$ by $\epsilon$, and let $\TT C[-]:A \to \B{nat}$ indicate a context; knowing that the interpretation of $\TT C$ is $k$-Lipschitz-continuous on some open set containing both $\model M$ and $\model N$, allows us to immediately deduce that $\TT C[M]$ improves on $\TT C[N]$ by $k \epsilon$. 
Observe that this will typically be the case when the Taylor expansions of $\TT C[M]$ and $\TT C[N]$ actually yields a \emph{finite} sum of at most $k$ terms, i.e.~when 
\begin{align}
\TT C[M] = \sum_{i=0}^{k} \TT D^{(k)}\Big[\lambda x.\TT C[x]\Big]\cdot M^{k}
\end{align}
and similarly for $\TT C[N]$. It is here worth recalling that, for $\STDLC$, a well-known result \cite{difflambda} is that the Taylor expansion of a closed application $MN$ is always finite, although its non-zero coefficients may be arbitrarily high. 
Notably, these observations suggest to study tropical versions of \emph{finiteness spaces} \cite{Ehrhard2005}, 
a variant of the relational semantics modeling strongly normalizing programs via \emph{finite} power series.
We mention this point in the Section~\ref{sec:concl}

@@

\begin{fact}
 By using \cite{Manzo2013}, it can be shown that $\LREL_!$ is a model of such calculus $\STLC_\oplus$. {\color{red}GIUSTIFICARLO BENE ?}
\end{fact}

@@

More formally, differential privacy is defined as follows \cite{Reed2010}:
\begin{definition}
Let $f: \mathsf{db}\to [0,1]^{X}$ and $\epsilon \in \BB R_{\geq 0}$. $f$ is said \emph{$\epsilon$-differentially private} if for all $x,x'\in \mathsf{db}$
differing by $L$ items, for all $s\in X$, 
$$
f(x)(s) \ = \ e^{\epsilon L} f(x')(s)
$$
\end{definition}

A well-studied approach to ensure differential privacy for higher-order programs is to use bounded type systems like $\mathsf{Fuzz}$ \cite{Reed2010}. Indeed, such systems come with an \emph{a priori} warrant that well-typed programs are Lipschitz.

The use of tropical semantics suggests how resource analysis could also be used to provide bounds for differential privacy. 
Suppose our probabilistic query $f$ can be expressed as a power series (this is what happens e.g.~in \emph{probabilistic coherent spaces} \cite{Ehrhard2011}). Then, if we discover, either by studying differential properties of $f$, or using methods from convex analysis as suggested in the previous section (e.g.~Theorem \ref{??}), 
 that the tropicalization $\trop f$ satisfies a Lipschitz condition, we may use this fact to deduce that $f$ is differentially private, as shown by the result below.

### 2023-01-16 section6.tex
commit

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{}. 
This correspondence, recalled in this section, and the fact that it leads to the definition of a linear $\Lawv$-category (i.e.~a model of the linear $\lambda$-calculus) of genealized metric spaces extending $\LREL$, is mostly a \emph{collage} of  \emph{folklore} results with more abstract results contained in recent literature in enriched category theory \cite{}.
Beyond this reconstruction, our main contribution is to show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by a suitable generalization of the construction of the exponential comonad of $\LREL$.

@@

Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
\begin{proposition}
$\Mod$ is a linear $\Lawv$-category.
\end{proposition}
\begin{proof}[Proof sketch]
The hom-sets $\Hom(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ \cite{} can be defined as the quotient of the usual tensor product of sup-lattices (see e.g.~\cite{}), under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$.
Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.

@@

Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$, satisfying 
\begin{align}
0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
\end{align}
This structure is often referred as a \emph{generalized metric space} \cite{}, or as a \emph{quasi-metric space} \cite{}. 
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.

A $\Lawv$-category is \emph{skeletal} \cite{} when 
$X(x,y)=0$ implies $x=y$, and 
 \emph{symmetric} when it coincides with its opposite category $X^{\op}(x,y):=X(y,x)$, i.e.~when $X(x,y)=X(y,x)$. 
 A metric space, in the usual sense, is thus the same as a skeletal and symmetric $\Lawv$-category.
  Notice that any $\Lawv$-category $X$ induces a skeletal category $X^{\sk}$ by quotienting points under $X(x,y)=0$, and a symmetric one by letting $X^{\sym}(x,y)=\max\{X(x,y),X^{\op}(x,y)\}$.

@@

Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973}. Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{}).

@@

First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies, Weight}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting the fact that tensors and products commute in $\Mod$, we define $!$ as the product of the symmetric algebras $!_{n}$.

@@

Using the fact that $\prod_{i}M_{i}\otimes  N\simeq (\prod_{i}M_{i})\otimes N$ holds in $\Mod$ [\cite{}, p.~???], we obtain the following:
\begin{theorem}
The functor $!M:= \prod_{n\in \BB N}!_{n}M$ is a Lafont exponential in $\Mod$.
Hence, $\Mod_{!}$ (equivalently, $\GMet_{!}$) is cartesian closed. 
\end{theorem}

@@

\item $\Mod$ can be equipped with a differential operator defined, for $f:!M\to N$, by 
\begin{align}\label{eq:dermod}
\Der[f](\alpha)=
\bigvee\left\{
f(\beta\cup [x]) \ \Big \vert  \ 
\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
\right\}
\end{align}
where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{}, 
satisfying all axioms (D1)-(D7) and (DCurry), see the Appendix.
\end{enumerate}

>>>>

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}. 
This correspondence, recalled in this section, and the fact that it leads to the definition of a linear $\Lawv$-category (i.e.~a model of the linear $\lambda$-calculus) of genealized metric spaces extending $\LREL$, is mostly a \emph{collage} of  \emph{folklore} results with more abstract results contained in recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}.
Beyond this reconstruction, our main contribution is to show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by a suitable generalization of the construction of the exponential comonad of $\LREL$.

@@

Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
\begin{proposition}
$\Mod$ is a linear $\Lawv$-category.
\end{proposition}
\begin{proof}[Proof sketch]
The hom-sets $\Hom(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices (see e.g.~\cite{Russo2007}), under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$.
Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.

@@

Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$, satisfying 
\begin{align}
0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
\end{align}
This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}, or as a \emph{quasi-metric space}. 
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.

A $\Lawv$-category is \emph{skeletal} \cite{Stubbe2014} when 
$X(x,y)=0$ implies $x=y$, and 
 \emph{symmetric} when it coincides with its opposite category $X^{\op}(x,y):=X(y,x)$, i.e.~when $X(x,y)=X(y,x)$. 
 A metric space, in the usual sense, is thus the same as a skeletal and symmetric $\Lawv$-category.
  Notice that any $\Lawv$-category $X$ induces a skeletal category $X^{\sk}$ by quotienting points under $X(x,y)=0$, and a symmetric one by letting $X^{\sym}(x,y)=\max\{X(x,y),X^{\op}(x,y)\}$.

@@

Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973}. Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{Willerton2013, Rutten1998}).

@@

First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting the fact that tensors and products commute in $\Mod$, we define $!$ as the product of the symmetric algebras $!_{n}$.

@@

Using the fact that $\prod_{i}M_{i}\otimes  N\simeq (\prod_{i}M_{i})\otimes N$ holds in $\Mod$ (see e.g.~\cite{Russo2007}), we obtain the following:
\begin{theorem}
The functor $!M:= \prod_{n\in \BB N}!_{n}M$ is a Lafont exponential in $\Mod$.
Hence, $\Mod_{!}$ (equivalently, $\GMet_{!}$) is cartesian closed. 
\end{theorem}

@@

\item $\Mod$ can be equipped with a differential operator defined, for $f:!M\to N$, by 
\begin{align}\label{eq:dermod}
\Der[f](\alpha)=
\bigvee\left\{
f(\beta\cup [x]) \ \Big \vert  \ 
\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
\right\}
\end{align}
where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}, 
satisfying all axioms (D1)-(D7) and (DCurry), see the Appendix.
\end{enumerate}

### 2023-01-16 section7.tex
commit

- correspondence between $\Lawv$-modules and $\Lawv$-categories.

- exponential structure of $\LCAT$.

- $\LCAT_{!}$ is a cartesian closed differential category

>>>>

The connections between the differential $\lambda$-calculus (and differential linear logic), the relational semantics, and the theory of non-idempotent intersection types is very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
As we said, the relational semantics over the tropical semi-ring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\B{PCF}$-like language with non-deterministic choice. 
\emph{Probabilistic coherent spaces} \cite{Ehrhard2011}, a variant of  the relational semantics, provide an interpretation of higher-order probabilistic programs
as analytic functions. In \cite{Ehrhard2022} it was observed that such functions satisfy a local Lipschitz condition somehow reminiscent of our examples in Section \ref{section4}.

The study of linear or bounded type systems for sensitivity analysis was initiated in \cite{Girard92tcs} and later developed \cite{Schopp, SchoppDalLago, Reed2010}.
As recalled in the paper, the use bounded exponentials ensures that well-typed programs satisfy a Lipschitz condition.
Related approaches, although not based on metrics, are provided by \emph{differential logical relations} \cite{dallago} and \emph{change action} models \cite{Picallo2019}.

More generally, the literature on program metrics in denotational semantics is vast. Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics. 
While standard categories of metric spaces are not models of the full simply typed $\lambda$-calculus, several constructions of cartesian closed categories of metric spaces can be found in the literature. For instance, 
\emph{ultra}-metric spaces form a CCC, and have been shown to model $\B{PCF}$ \cite{Escardo1999}.
Also \emph{partial} metrics, introduced in \cite{matthews}, have been shown to provide models of $\STLC$, under suitable generalizations \cite{Geoffroy2020}.
More generally, \cite{Clementino2006} provides a general characterization of exponentiable objects in categories of (generalized) metric spaces, and \cite{PistoneLICS, PistoneFSCD2022} provide other ways to construct CCCs of
(generalized) metrics, including one based on locally Lipschitz maps, 
using ideas from {differential logical relations} and quantitative equational theories \cite{Plotk}.

Motivated by connections with computer science and fuzzy set-theory, 
the abstract study of generalized metric spaces in the framework of \emph{quantale}- or even \emph{quantaloid}-enriched categories has led to a vast literature in recent years \cite{Hofmann2014, Stubbe2014}, 
and its connections with tropical mathematics have been explored e.g.~in \cite{Fuji, Willerton2013}. Moreover, applications of quantale-modules to both logic and computer science have also been studied \cite{Abramsky1993b, Russo2007}.

Connections between program metrics and the differential $\lambda$-calculus have been already suggested in \cite{PistoneLICS}; moreover, \emph{cartesian difference categories} \cite{Picallo2020} have been proposed as a way to relate derivatives in differential categories with those found in change action models.

Finally, applications of tropical mathematics in computer science abound, originally for automata theory \cite{Chua1992, Simon}, and more recently
 in optimization methods for machine learning and other statistical models (see e.g.~\cite{Maragos2021, Pachter2004, Zhang2018}), optimization \cite{Akian2011, Akian2012}, and convex analysis \cite{Lucet2009}. While our discussion in Section \ref{section5} is inspired by a well-known application of tropical polynomials to Hidden Markov Models \cite{Pachter2004},  
the vast literature in this domain lets us think that other ways to 
apply tropical semantics to the analysis of higher-order programs might be studied.

%
%Other connections tropical/metric -> Fuji, ??
%Applications of tropical to computer science.
%Log-probabilities.
%Quantale-modules -> Abramski?
%
%
